---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r, setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(brech)
set.seed(42)

```

## Introduction

To create a model simulation we construct an R S3 object using the `stochastic_model` class. This expects a list of reactions including the `transitions`: the impact on the population and `rate`: the rate at which the event takes place. e.g. the first reaction represents an infection event that occurs at rate $\beta S I$ where $S$ is the population of susceptible, $I$ is the population of infected and $\beta$ is the population infectivity.

```{r}
#| echo: true
#| message: true
reactions <- list(
  infection = list(
    transition = c(S = -1, E = +1),
    rate = function(x, p, t) p$beta * x["S"] * x["I"]
  ),
  incubation = list(
    transition = c(E = -1, I = +1),
    rate = function(x, p, t) p$theta * x["E"]
  ),
  recovery = list(
    transition = c(I = -1, R = +1),
    rate = function(x, p, t) p$gamma * x["I"]
  ),
  case_detection = list(
    transition = c(C = +1),
    rate = function(x, p, t) p$case_rate * x["I"]
  ),
  vaccinate = list(
    transition = c(V = +1, S = -1),
    rate = function(x, p, t) 0 * x["S"]
  )
)
sim_scenario <- load_model_params() |>
  vaccinate_initial_conditions(0.8) |>
  update_parameters(beta=0.02)

sm <- stochastic_model(reactions,sim_scenario)
sm

```

To run with default settings, just use the `run` function

```{r}
#| echo: false
r <- run(sm)
r |> ggplot(aes(x=time,y=C)) + geom_line()
```

## Model fitting

Approximate Bayesian Computation (ABC) is used to perform model fitting. This is done by defining a summary statistic and then running the summary statistic on the data. We will use the data above as an example and create a summary statistic for the cumulative number of cases after 10 days.

We then draw priors from a defined distribution for each parameter to be fit. In the example below the parameter `beta` if fit to the cumulative number of cases.

```{r}
# generate summary statistics from run
# These are the number of cases on a given day
daily_cases <- sum(get_daily_cases(r)$daily_incidence)

# simulate from prior
priors <- tibble("beta" = pmax(0,rnorm(200,mean=0.1,sd=0.1)))

# statistical summary of simulation
stat_func <- function(m){
  #' get total cases 
  m |>
    get_daily_cases() |>
    dplyr::summarise(total_cases=sum(daily_incidence))
}

res <- abc_stochastic_model(sm,priors,daily_cases,stat_func)

plot(res,"beta")
```

We can similarly fit to daily cases. Note that we need to exclude the first few days in the simulation where the variance in case incidence is too low. Fitting the incidence appears to improve the model fit with this small sample.

```{r}
# simulate from prior
priors <- tibble("beta" = pmax(0,rnorm(200,mean=0.1,sd=0.1)))

# statistical summary of simulation
stat_func <- function(m){
  #' get total cases 
  m |> 
    get_daily_cases() |> 
    filter(day > 3) |> 
    pivot_wider(names_from = "day",values_from="daily_incidence", 
                names_prefix = "day_")
}

daily_cases <- stat_func(r)

res <- abc_stochastic_model(sm,priors,daily_cases,stat_func)

plot(res,"beta")
```

The fit also provides the final state of each simulation from the posterior which we can compare to our fitted data

```{r}
data.frame(r) |>
  filter(time == max(time)) |>
  dplyr::select(-time) |>
  mutate(type = "true value") |>
  bind_rows(
    mutate(res$state,type="posterior")
  ) |>
  pivot_longer(-type,names_to = "state", values_to = "value") |>
  ggplot(aes(x=state,y=value,color=type)) +
  geom_jitter(alpha=0.5)

```

## Scenario projection

We can then perform projections using a `abc_stochastic_model` object. The projection will sample over the final state of the model in the posterior as well as any associated parameters and inherit other parameters from the underlying `stochastic_model` class. We project forward in time using the `projection_stochastic_model` function which also comes with default plotting behaviour.

```{r}
projections <- projection_stochastic_model(res)
plot(projections, state = "I")
plot(projections, state = "C")
```

The `projection_stochastic_model` function can also incorporate scenarios directly without the need for model fitting. This is useful if there is a need to project under different scenarios where there is no current data to inform the model fit. Under this scenario we draw from a defined distribution of $\beta$ parameters

```{r}
parameters <- tibble("beta" = runif(200,min=0.08,max=0.12))
scenarios <- scenario_stochastic_model(sm,parameters)
projections <- projection_stochastic_model(scenarios)
plot(projections, state = "I")
```

A flow chart of the dependency between defined S3 classes in the package is provided below. The base class is `stochastic_model` which defines the model events, rates, parameters and initial states. A `summary` of the object can be provided and a simulation can be drawn from the model using the function `run()`. Model fitting can be performed using the `abc_stochastic_model` or alternatively a scenario can be generated using the `scenario_stochastic_model` class. Finally a projection can be generated from either a scenario or model fit using `projection_stochastic_model`.

```{mermaid}
flowchart TD
  M("load_parameters()") --> A
  R("reactions") --> A
  A[stochastic_model] --> B[abc_stochastic_model]
  A --> C[scenario_stochastic_model]
  B --> D[projection_stochastic_model]
  C --> D
```

## Age-structured model

For an age structured model we need to set-up reactions for each age-group. This can be done explicitly as above, however would quickly become cumbersome for a large number of age-groups. Instead we can construct this by looping over age groups and defining each of the reactions e.g. "infection" and "recovery". This requires that the rates functions are built inside a function factory to ensure that the correct age-parameters are provided to each function. For more information see [this chapter](https://adv-r.hadley.nz/function-factories.html).

First we construct a simple contact age contact matrix with two age groups,

```{r}
age_groups <- c("<18", "18+")
contact_matrix <- matrix(c(20,5,5,10), nrow = 2)
rownames(contact_matrix) <- age_groups
colnames(contact_matrix) <- age_groups
print(contact_matrix)
```

Next we use the `age_groups` and `contact_matrix` to populate the `reactions` list. To illustrate the code below provides a simple age-structured SIR model.

```{r}
reactions <- list()
  for (i in seq_along(age_groups)){
        age_group <- age_groups[i]
        ###########################
        #  infection transitions  #
        ###########################
        transition <- function() {
          trans <- c()
          trans[paste0("S_",age_group)] <- -1
          trans[paste0("I_",age_group)] <- +1
          trans
        }
        rate <- 
          function(i,age_group){
            force(i)
            force(age_group)
            function(x, p, t) {
              I <- x[paste0("I_", age_groups)]
              S <- x[paste0("S_", age_group)] 
              lambda <- p$beta * S * sum(contact_matrix[,i] * (I / p$N))
              lambda  
            }
          }
        reactions[[paste0("infection_",age_group)]] <- list(
          "transition"=transition(),
          "rate" = rate(i,age_group)
        )
        ##############################
        #    recovery transitions    #
        ############################## 
        transition <- function() {
          trans <- c()
          trans[paste0("I_",age_group)] <- -1
          trans[paste0("R_",age_group)] <- + 1
          trans
        }
        rate <- 
        function(age_group){
          force(age_group)
          function(x, p, t) {
            I <- x[paste0("I_", age_group)]
            p$gamma * I
          }
        }
        reactions[[paste0("recovery_",age_group)]] <- list(
          "transition"=transition(),
          "rate" = rate(age_group)
        )
      
    }
```

We then apply these reactions to construct a `stochastic_model` object. We use the function `create_age_initial_conditions` to convert the initial conditions from a vector format to the format required for the object.

```{r}
age_scenario <- list(
  params = list(beta = 0.09, gamma = 0.2, N = 1000),
  initial_states = list(
    S = c(200, 800),
    V = c(0, 0),
    I = c(1, 0),
    R = c(0, 0)                    
    ),
  sim_args = list(T = 10)
)

age_scenario <- age_scenario |>
  create_age_initial_conditions(age_groups)


sm <- stochastic_model(reactions,age_scenario)
print(sm)
```

As before we can run this simulation and plot

```{r}
r <- sm |>
    brech::run() |>
    as.data.frame() |>
    tidyr::pivot_longer(
      cols = tidyselect::starts_with("I_"),
      names_to = "age_group",
      names_prefix = "I_",
      values_to = "I"
    )
r |> ggplot(aes(x=time,y=I, color= age_group)) + geom_line()
```

We can also vaccinate the initial population by age group. In the example below we imagine that the \<18 age group is 10% vaccinated and the 18+ age group is 90% vaccinated. Notice that the initial values are now updated.

```{r}

sm |> vaccinate_initial_conditions(c("<18"=0.1,"18+"=0.9))
```

We can similarly generate projections using the `scenario_stochastic_model` object,

```{r}

parameters <- tibble("beta" = runif(200,min=0.08,max=0.12))
scenarios <- scenario_stochastic_model(sm,parameters)
projections <- projection_stochastic_model(scenarios)
plot(projections, state = "I")

```

A series of helper functions can also be used to construct quantiles and plot summaries from an age-structured model as long as the states are named in the format `{model_state}_{age_group}`.

```{r}
projections |> 
  projection_quantiles_by_age_group("I") |>
  plot_projections_by_age_group("I")
```

The final state can also be extracted by first making the projections longer and then examining the recovered individuals at the max time,

```{r}
projections$projection |>
  create_age_group_column("R") |>
  filter(time == max(time)) |>
  ggplot(aes(fill=age_group,x=R)) +
  geom_density(alpha=0.2) +
  labs(x = "Final size",fill="")
```

## Fitting initial conditions

`abc_stochastic_model` can also take in initial states in its fitting using the optional `states` argument. This needs to be a complete model state for each row e.g.,

```{r}
N <- c(200,800)
vaccine_rate <- c(0.1, 0.5)
infection_rate <- c(0.02,0.01)
nsamples <- 100
V <- matrix(rbinom(2*nsamples, size = rep(N, times = nsamples), 
              prob = rep(vaccine_rate, times = nsamples)),
       nrow = nsamples, byrow = TRUE)
S <- matrix(rep(N,times=nsamples), nrow=nsamples,byrow = TRUE) - V
# simulation initial infections 
I <- matrix(rbinom(2*nsamples, size = rep(N, times = nsamples), 
              prob = rep(infection_rate, times = nsamples)),
       nrow = nsamples, byrow = TRUE)
zero_state <- matrix(rep(0,times=2*nsamples), nrow=nsamples)
states <- as.data.frame(cbind(S, V, I, zero_state))
colnames(states) <- c("S_<18", "S_18+", "V_<18", "V_18+", 
                      "I_<18", "I_18+", "R_<18", "R_18+")
head(states)
```

We can also project from a set of states with associated parameters

```{r}
# fix beta parameter
sm <- sm |> update_parameters(beta = 0.1)

# define scenario based on list of states
example_scenario <- scenario_stochastic_model(sm, states = states)
state_projections <- projection_stochastic_model(example_scenario)
plot(state_projections, "I")
```

We can perform model fitting as before using the `states` as a prior in the model that it is sampled from. In the example below we generate some data with an initial distribution of vaccinated among children and adults and use the number of recovered in each age group as the summary function.

```{r}
age_scenario <- list(
  params = list(beta = 0.1, gamma = 0.2, N = 1000),
  initial_states = list(
    S = c(190, 400),
    V = c(10, 400),
    I = c(1, 0),
    R = c(0, 0)                    
    ),
  sim_args = list(T = 10)
)

age_scenario <- age_scenario |>
  create_age_initial_conditions(age_groups)


sm <- stochastic_model(reactions,age_scenario)

r <- brech::run(sm) |> as.data.frame()

stat_func <- function(m){
  #' get total cases by age group
  m[nrow(m),startsWith(colnames(m),"R")]
}

cumulative_cases <- stat_func(r)

```

```{r}
# simulate from prior
priors <- tibble("beta" = pmax(0,rnorm(100,mean=0.1,sd=0.1)))


res <- abc_stochastic_model(sm,priors,cumulative_cases,stat_func,
                            states = states)
```

## Using `multisession` and `progressr`

